First and foremost, I express my profound gratitude to God for the gift of life, grace and opportunity bestowed onto me for bringing me this far in life and helping me complete another step towards achieving my dreams.\comm{You used \textbackslash\textbackslash[5pt] at the end of this paragraph.  Please just separate paragraphs with blank lines here and in the rest of the thesis.  In general \textbackslash\textbackslash is best avoided outside of tabular or array environments.  It sort of works, but it's not a normal way to mark a paragraph in \LaTeX\ and it can cause worse output with underfull hboxes.  I don't see any reason for the [5pt] either, so I'd also remove that.  Using \textbackslash par is also not recommended.  Finally, it's good practice to put some blank lines around the \textbackslash section, etc.  It makes them easier to spot in the \LaTeX\ code.}

I would also like to express my deepest gratitude to my thesis advisor, Professor Michael Stewart for his unwavering support, guidance and impact on the completion of this thesis. I \rep{had}{have} absolutely zero knowledge or idea on this subject before starting this thesis, but his expertise, patience and insightful feedback have been invaluable in shaping this thesis and my growth as a mathematician. It is such a great privilege to have had the opportunity to learn from you. God bless you sir.

Additionally, I would also like to extend my sincere appreciation to the members of my committee, Professor \rep{Russell Jeter}{Jeter Russell}, and Professor Vladimir Bodarenko, for their time, thoughtful suggestions and feedback on this work. Your work have greatly influenced my research and your perspectives have enriched this thesis and helped me refine my ideas.\\[5pt]
To my colleagues and friends in the department \textemdash \comm{Use a textemdash here instead of -.} John Ajayi, Xavier Sodjavi, Sheriff Akeeb, Akinwale Famotire, Emeka Mazi, to mention a few, I say a big thank you for creating a highly simulating and collaborative environment for learning. My sincere gratitude also goes to the faculty members of the Department of Mathematics at Georgia State University, many of whom I have had the honor of learning from. I am particularly thankful to Dr. Zhongshan Li, Professor Alexandra Smirnova, and Professor Mariana Montiel, to mention a few, for their mentorship, expertise, and encouragement throughout my academic journey. Their dedication to teaching and research has been a constant source of inspiration. Additionally, I extend my thanks to the entire staff of the department for their support and assistance, which have been instrumental in creating a conducive environment for learning and research.\\[5pt]
Finally, I would like to acknowledge my parents, Mr. and Mrs. Adebesin, my siblings and several father and mother figures in my life. Mr. and Mrs. Olawuyi, Mrs. Abioye, Mr. Agboola for their unwavering support. \comm{This was all a very nice acknowledgment.}
